---
title: "MCMC in Stan"
author: "STAT 341, Spring 2023"
date: "2023-03-20"
output: 
  html_document:
    toc: true
    toc_float: true
    code_download: true
---

```{r setup, include=FALSE}
library(tidyverse)
library(tidybayes)
library(rethinking)
library(ggformula)
library(rstan)
library(bayesplot)
library(faraway)
knitr::opts_chunk$set(echo = TRUE)
options(mc.cores = parallel::detectCores())
theme_set(theme_minimal(base_size = 18))
```

## Data

The data for our example model is presented in Chapters 8-9 of *Statistical Rethinking,* on gross domestic products and ruggedness of terrain in countries around the globe. 

The code below loads and prepares the dataset:

```{r}
# rugged dataset is in the rethinking package
data("rugged")
rugged <- rugged |>
  mutate(
    # convert GDP data to natural log scale
    log_gdp = log(rgdppc_2000),
    # standardize log(GDP) by dividing by its mean value (see CH. 8 for rationale)
    log_gdp_std = log_gdp / mean(log_gdp, na.rm = TRUE),
    # standardize terrain ruggedness by dividing by its max value
    # (not z-score b/c want to keep 0 ruggedness to mean totally flat)
    rugged_std = rugged / max(rugged, na.rm = TRUE),
    # create categorical variable for Africa or Not Africa
    region = ifelse(cont_africa == 1, 'Africa', 'Not Africa')) |>
  drop_na(country, log_gdp_std, rugged_std, region) |>
  select(country, log_gdp_std, rugged_std, region)
# prepare data for use in Stan
stan_data <- compose_data(rugged)
nrow(rugged)
```

A note about `rugged_std`: we didn't want to center this variable since a ruggedness value of 0 means totally flat land, and we want to preserve that meaning. However, if we *don't* center, then defining priors for the intercept for any model with a ruggedness predictor will be harder. So you will see that in the model we use `rugged_std - 0.215` as a predictor -- 0.215 is the mean value of `rugged_std` and the idea is that setting up the model this way kind of lets us have both: a meaning for `rugged_std = 0` *and* a prior for the intercept that can be interpreted as the "average overall response value."

## Fitting with Stan

You can save a Stan program in a dedicated flat text file, or a string in R (shown below). It has 3 parts:

- data: how many observations are there, and what are the variables?
- parameters: what parameters will the model produce posteriors for?
- model: how is the likelihood defined, and what are the priors?

```{r, stan-ruggedness}
stan_program <- '
data {
  int<lower=1> n;     // number of observations
  vector[n] log_gdp_std;    // response
  vector[n] rugged_std;     // predictor
  array[n] int region;            // categorical predictor, africa or not
}
parameters {
  real<lower=0> sigma; // std of response, single continuous value
  vector[2] a; // vector of two numeric values: intercepts for Africa and Not Africa
  vector[2] b; // vector of two numeric values: slopes for Africa and Not Africa
}
model {
  vector[n] mu; // vector of n values: expected log_gdp_std for each observation
  for (i in 1:n) { // loop over the n cases in the dataset to estimate mu_i values
    mu[i] = a[region[i]] + b[region[i]] * (rugged_std[i] - 0.215); // expected log_gdp_std is region-specific intercept + region-specific slope * (rugged_std = 0.215)
  }
  a ~ normal(1, 0.1); // prior for both intercepts
  b ~ normal(0, 0.3); // prior for both slopes
  sigma ~ exponential(1); // prior for sigma
  log_gdp_std ~ normal(mu, sigma); // defining likelihood in terms of mus and sigma
}
'

# fit the model defined above to the stan_data on ruggedness and GDP
rugged_model <- stan(model_code = stan_program, data = stan_data)
```

We will learn more about how to "tell" Stan things in Stan programs as we go. But for now, note 2 very useful reference resources:

- [Stan code for models fitted in *Statistical Rethinking*](https://vincentarelbundock.github.io/rethinking2/) by Vincent Arel-Bundock
- Stan reference for [available continuous distributions](https://mc-stan.org/docs/functions-reference/continuous-distributions.html) (and the full [Stan functions reference](https://mc-stan.org/docs/functions-reference/index.html))

## Viewing Results

```{r, stan-summary}
rugged_model
```

```{r, stan-trace}
mcmc_trace(rugged_model)
```

## Posterior Sample

To get the post-warmup samples from the posterior, in the form of a data frame, from a fitted Stan model in R:

```{r, post-samp-stan}
psamp <- as.data.frame(rugged_model)
glimpse(psamp)
```

- What is `lp__`? It's a measure of the log-likelihood of the data given the sampled parameter values. ("Up to a constant," meaning the values can be compared to one another, but are only proportional to the likelihood.)

We might want to rename the variables in the posterior sample for easier use in R (and easier remembering-what-they-are):

```{r}
psamp <- psamp |>
  rename(intercept_Africa = `a[1]`,
         intercept_NotAfrica = `a[2]`,
         slope_Africa = `b[1]`,
         slope_NotAfrica = `b[2]`,
         relative_log_lik = lp__)
glimpse(psamp)
```

## Your Turn

Convert a model you have previously fitted (or one from any previous class example) to be fitted using Stan.
```{r}
data(fat)
head(fat)
```

```{r}
fat_data <- fat |>
  mutate(
    density_std = (density - mean(density)) / sd(density),
    brozek_std = (brozek - mean(brozek)) / sd(brozek)
  ) |> 
  drop_na(density, brozek) |>
  select(density, brozek)

# prepare data for use in Stan
stan_data <- compose_data(fat_data)
nrow(fat_data)
```

```{r, stan-ruggedness}
stan_program <- '
data {
  int<lower=1> n;     // number of observations
  vector[n] brozek_std;    // response
  vector[n] density_std;     // predictor
}
parameters {
  real beta_0; 
  real beta_1; 
  real<lower=0> sigma; 
}
model {
  vector[n] mu; // vector of n values: expected log_gdp_std for each observation
  for (i in 1:n) { // loop over the n cases in the dataset to estimate mu_i values
    mu[i] = beta_0 + beta_1 * density_std[i];
  }
  beta_0 ~ normal(0, 10);
  beta_1 ~ normal(0, 10);
  sigma ~ uniform(0, 10);
  brozek_std ~ normal(mu, sigma);
}
'

# fit the model defined above to the stan_data on density and brozek
fat_model <- stan(model_code = stan_program, data = stan_data)
```

```{r, stan-summary}
fat_model
```

```{r, stan-trace}
mcmc_trace(fat_model)
```

```{r, post-samp-stan}
psamp <- as.data.frame(fat_model)
glimpse(psamp)
```


```{r}
psamp <- psamp |>
  rename(intercept = beta_0,
         slope = beta_1,
         sigma = sigma,
         relative_log_lik = lp__)
glimpse(psamp)
```

- Can you get it running?
- Is the posterior similar to what you got with grid search or `quap()`?